Fourier-domain modulations and delays of gravitational-wave signals

### 2017-09-11 docs/paper_method.tex
I rearranged the introductory part as discussed and made other minor edits.

\newcommand{\SM}[1]{{\color{Red} #1}}

@@

\author{Sylvain Marsat}
\affiliation{Max Planck Institute for Gravitational Physics (Albert Einstein Institute), Am M\"uhlenberg 1, Potsdam-Golm, 14476, Germany}
\author{John G. Baker}
\affiliation{Gravitational Astrophysics Laboratory, NASA Goddard Space Flight Center, 8800 Greenbelt Rd., Greenbelt, MD 20771, USA}

@@

We present a Fourier-domain based approach modulations and delays of gravitational wave signals, a problem which arises in two different contexts. For space-based detectors like LISA, the orbital motion of the detector introduces a time-dependency in the response of the detector, consisting of both a modulation and a varying delay. In the context of signals from precessing spinning binary systems, a useful tool for building models of the waveform consists in representing the signal as a time-dependent rotation of a quasi-non-precessing waveform. In both cases, being able to compute transfer functions for these effects directly in the Fourier domain would allow to gain performance for data analysis applications by using fast frequency-domain waveforms. Our results recover previous approaches based on an extension of the stationary phase approximation, generalize them by including delays and higher-order corrections, and can in principle be applied to full inspiral-merger-ringdown signals. In the LISA case, we find that higher-order corrections remain small except at the two extremes of the detector's frequency band, where we show that including them reduces the errors. In the case of precessing binaries, we perform a limited exploration of the merger-ringdown range based on a toy model for the precession. We find that, depending on the post-merger frame velocity, higher-order corrections can become quantitatively large to the point of challenging our perturbative formalism, while affecting mainly the amplitude and giving raise to a limited unfaithfulness. We develop an alternative direct convolution approach to represent accurately these post-merger features.

@@

With the recent detections of gravitational-wave signals emitted by the coalescences of binary black holes, announced by the LVC collaboration after the first obervation run of the two LIGO instruments~\cite{LIGO-theevent-2016,LIGO-christmasevent-2016,LIGO-O1BBH-2016}, gravitational-wave astronomy has entered its observational era. With the recent start of LIGO's second observation run, and later the extension of the ground-based detectors network with Virgo~\cite{Virgo}, KAGRA~\cite{KAGRA} and LIGO-India~\cite{INDIGO}, many more detections from compact object coalescences are to come, with binary black holes being expected to be the dominant sources~\cite{} \SM{[C]}. Moreover, the European Space Agency has issued a mission call for its L3 space mission with a tentative launch around 2034, adressing the science theme ``The Gravitational Universe'', possibly opening the way for a space-borne observatory based on the concept of the Laser Interferometer Space Antenna (LISA)~\cite{LISA09,LISA17}. Among other sources of gravitational waves, LISA will be able to detect and characterize comparable-mass binary black hole coalescences to cosmological distances over a wide range of masses, from high-redshift observations of supermassive black hole binaries with $M\sim 10^{7} \Msol$ down to the observation of LIGO-type sources with $M\sim 10^{2} \Msol$~\cite{Sesana16}.

Data analysis for gravitational-wave observations of compact binary coalescences require accurate models (or templates) for the signals, both for ensuring efficient detections of signals that may be buried in instrumental noise, and to extract the physical parameters of the source in a subsequent analysis. This has prompted an important and ongoing effort to improve gravitational wave templates. Two important qualitative features for the improvement of the templates are the coverage of all phases of the evolution of the system, including the inspiral, merger and ringdown (IMR), which became possible thanks to the breakthrough of numerical relativity~\cite{Pretorius05,Baker+06,Campanelli+06}, and the inclusion of the effects of the spins of the compact objects, that can lead to a precession of the orbital plane~\cite{ACST94,K95}. Besides accuracy, the computational efficiency of template generation is crucial, as Bayesian analysis for parameter estimation of gravitational-wave signals, as was performed for the LIGO detections~\cite{LIGO-theeventPE-2016,LIGO-O1BBH-2016}, may require millions of evaluations of the likelihood function to sample the posterior probability distribution.

State-of-the-art IMR templates need to combine information from the perturbative results of post-Newtonian theory covering the inspiral (see e.g.~\cite{BlanchetLiving}) and from numerical relativity simulations covering the end of the inspiral and the merger-ringdown phase (see e.g.~\cite{Pfeiffer12}). The Effective-One-Body (EOB) approach~\cite{BD99} does so by incorporating a resummation of post-Newtonian results and a calibration to numerical relativity, and has been extended to include both spins of the two bodies~\cite{Taracchini+13, Pan+13, Bohe+16}. Phenomenological templates extend post-Newtonian models by postulating an analytic ansatz for the Fourier-domain amplitude and phase, fitting coefficients against numerical relativity waveforms~\cite{Husa+15, Khan+15}, and can include a simplified model for precession~\cite{Hannam+13}. The second type of models produce directly the amplitude and the phase of the Fourier-domain waveform in the form of a piecewise analytic function. EOB waveforms, by contrast, require to integrate ordinary differential equations in the time-domain, which can be time-consuming. For the non-precessing case, Reduced Order Models (ROM), also called surrogate models, have been developed to speed up considerably the waveform generation, without losing accuracy~\cite{Field+13, Puerrer14, Bohe+16}. Both approaches are therefore able to produce efficiently IMR Fourier-domain waveforms. An important aspect is that the resulting waveforms can be represented by an amplitude and phase for each mode, with only a few hundred samples~\cite{Puerrer14}.

A complete representation of the spin effects in fast IMR templates still remains a frontier of gravitational wave signal modelling. In presence of misaligned spins, the system will endure a precession of the orbital plane as it evolves, leading to modulations of the signal as seen by the observer~\cite{ACST94}. Precessing EOB templates incorporate all six parameters of the two spins~\cite{Pan+13, BTB16} but still require a costly time-domain integration; while precessing phenomenological templates~\cite{Hannam+13} decrease the dimensionnality by using effective spin parameters representing only the dominant effects of the spins. To model the effect of precession on the emitted waveform, it has been proposed~\cite{BCV03b, BCPTV05, Schmidt+10, OShaughnessy+11, Boyle+11} to decompose precessing waveforms by performing a time-dependent rotation, following the precession of the orbital plane. The resulting waveform in the rotated frame can then be modelled by a non-precessing waveform, an approximation which is used both in the construction of the inspiral part of precessing EOB waveforms~\cite{Pan+13} and in the construction of precessing phenomenological waveforms~\cite{Hannam+13}. To follow this modelling approach and create efficiently Fourier-domain waveforms, one needs to understand how to translate the time-domain modulation created by the frame rotation into a Fourier-domain transfer function, which is our first motivation.

Prospective parameter recovery studies for LISA observations of comparable-mass binary black hole coalescences aim at characterizing the scientific performance of the future experiment. They become a timely topic with the upcoming effort for the ESA L3 mission characterization in the next few years. However, the recent progresses in waveform modelling, most notably the availability of cheap IMR waveforms, have not yet made their way to the field of LISA data analysis studies. In the past, such studies used mainly post-Newtonian, inspiral-only models, and were limited to the Fisher-matrix formalism instead of full Bayesian analysis (among numerous works, see e.g. \cite{Cutler97,LISAPE09}). Refs.~\cite{McWilliams+09,McWilliams+10,McWilliams+11} explored the effects of including the merger and ringdown, keeping the mass parameters fixed. However, the Fisher matrix analysis, although considerably cheaper to perform, is valid only in the limit of a high signal to noise ratio (SNR) and has known shortcomings~\cite{Vallisneri08}. Several implementations of Bayesian analyses were developed~\cite{CP06,Wickham+06,Babak+08,GP09,Feroz+09,Petiteau+10,PC13} (see also~\cite{MLDC4} and references therein), with Ref.~\cite{PC15} proposing a direct comparison of Fisher and Bayes analyses, but they were limited to inspiral-only waveforms. One exception was Ref.~\cite{Babak+08} where IMR signals were used in a Bayesian study, but there the masses of the black holes were kept fixed. In the recent Ref.~\cite{Klein+15}, the limitation to inspiral-only precessing post-Newtonian waveforms was compensated by an SNR reweighting procedure. Thus, due mainly to their high computational cost, full-fledged Bayesian parameter estimation simulations for IMR signals are still missing.

By contrast with ground-based detectors, the response of a LISA-type instrument has the specificity of being time-dependent~\cite{Cutler97}. For signals possibly accumulating over several months, the motion and change of orientation of the detector constellation along its orbit lead to significant time variability in the form of a modulation and a varying delay, which contain information about the localization of the gravitational-wave source in the sky. This complicated detector response increases the computational cost of parameter-estimation analyses. To leverage the performance of state-of-the-art Fourier-domain IMR templates~\cite{BTB16,Khan+15}, one needs to be able to process the signals through the time-dependent response of the dectector while staying in the Fourier domain, which is our second motivation.

The purpose of this paper is to propose a formalism allowing to transpose directly in the Fourier domain the processing of a signal through a time-domain modulation and delay, while retaining the compactness of a Fourier-domain amplitude and phase representation of the signals. This will allow us to address both the issue of the Fourier-domain response of the LISA instrument, as well as the issue of Fourier-domain precession modulation for IMR signals from precessing binaries.

@@

The plan of the paper is as follows. In Sec.~\ref{sec:motivation}, we will introduce our notations and conventions for Fourier-domain waveforms, motivate the problem of Fourier-domain modulation and introduce the relevant timescales for both the response of LISA-type detectors and the modulation of precessing signals. In Sec.~\ref{sec:formalism}, we present our general formalism, give its leading order approximation as well as higher-order corrections, introduce new timescales based on the Fourier-domain signal, refine the previous results for both the quadratic-in-phase corrections and the treatment of the delays, and make the connection to previous approaches of the problem. We then apply our formalism to the response of the LISA detector in Sec.~\ref{sec:LISA}, and to the case of signals from precessing binaries in Sec~\ref{sec:precession}. We discuss and summarize our results in Sec.~\ref{sec:discussion}.

@@

\subsection{Notations and conventions}
\label{subsec:notations}

We recall here an approximation widely used to model gravitational waves signals emitted by inspiralling binaries, the Stationary Phase Approximation (thereafter SPA). It applies in general to chirping signals, and we refer the reader to~\ref{} for details. One first writes the time-domain signal in amplitude and phase form as $h(t) = a(t) e^{-2i\varphi(t)}$, where for gravitational wave signals $\varphi$ will correspond to the orbital phase, with the orbital frequency being $\omega = \dot{\varphi}$. In order to keep close to the notations used in the gravitational wave literature, we introduced a factor of 2 in the phase of the wave, which is appropriate for the dominant 22 harmonic of the signal. The approximation then applies to signals verifying the conditions \SM{[say that the third is sometimes missing ?]}
\be\label{eq:conditionSPA}
	\left| \frac{\dot{a}/a}{\omega} \right| \ll 1\,, \quad \left|\frac{\dot{\omega}}{\omega^{2}} \right| \ll 1\,, \quad \left| \frac{(\dot{a}/a)^{2}}{\dot{\omega}} \right| \ll 1 \,.
\ee
Since the integral~\eqref{eq:defFT} defining the Fourier transform is rapidly oscillatory unless the term $2\pi f t$ cancels the evolution of $-2\varphi(t)$, its support is well centered around the point of stationary phase. This determines the time-to-frequency correspondence in the stationary phase approximation, and leads to the definition of this time as an implicit function of frequency by the relation
\be\label{eq:deftfSPA}
	\omega(t_{f}^{\rm SPA}) = \pi  f \,.
\ee
For a chirping signal of increasing phase, $\omega>0$ and $\dot{\omega}>0$, there is a unique point of stationary phase, located in the positive frequency range $f>0$. Using the conditions above, one can formally expand the signal around $t_{f}^{\rm SPA}$ to quadratic order in time, according to
\begin{align}
	\tilde{h}_{\rm SPA} (f) &\simeq a(t_{f}^{\rm SPA}) \exp\left[2i\pi f t_{f}^{\rm SPA}-2i\varphi(t_{f}^{\rm SPA}) \right] \nn\\
	& \qquad \cdot \int \ud t \, e^{-i \dot{\omega} (t_{f}^{\rm SPA}) (t-t_{f}^{\rm SPA})^{2}} \,.
\end{align}
Note that to be able to treat the amplitude as a constant in the integral above, we used the third condition in~\eqref{eq:conditionSPA}. The resulting complex Gaussian integral yields\footnote{Note that the expressions given here are valid for the $22$ mode of the waveform.}
\begin{subequations}
\begin{align}
	\tilde{h}_{\rm SPA}(f) &= A_{\rm SPA}(f) e^{-i\Psi_{\rm SPA}(f)} \,, \\
	A_{\rm SPA}(f) &= a(t_{f}^{\rm SPA}) \sqrt{\frac{\pi}{\dot{\omega}(t_{f}^{\rm SPA})}} \,, \\
	\Psi_{\rm SPA}(f) &= 2\varphi(t_{f}^{\rm SPA}) - 2\pi f t_{f}^{\rm SPA} + \frac{\pi}{4} \,.
\end{align}
\end{subequations}
Ref.~\cite{Droz+99} evaluated the first correction to this approximation, within the context of post-Newtonian signals, and found that it can be considered as a term of the fifth post-Newtonian order, beyond the accuracy level of our current best models~\cite{BlanchetLiving}.

@@

The response of a detector of the LISA type to an incident gravitational wave can be written in two different, equivalent forms, in terms of phase or frequency measurements. Here we will work with the second representation, which will prove more convenient for our purposes. Moreover, various notations and conventions have been used in the literature to label the spacecrafts and describe their orbits. We refer the reader to~\cite{Vallisneri04} for a comparative account on these various conventions. In this work, we will use the conventions of~\cite{Vallisneri04}, which were also used in the Mock LISA Data Challenges (MLDC)~\cite{} \SM{[C]}.

@@

@@

@@

Let us consider, in general, the problem of determining the Fourier transform of a signal $h(t)$ to which we apply a varying delay $d(t)$ and a multiplication by a modulation function $F(t)$. In the case of precessing binaries, the delays will be absent, and the modulation functions will be the time-dependent Wigner coefficients applied to rotate the waveform from a precessing frame, where it verifies a restored mode hierarchy, to an inertial frame where the observations take place. In the case of a LISA-type detector, the signal $h(t)$ will simply be the waveform in a fixed heliocentric frame, and the delays will come from the motion of each detector in the wave front, while the modulation will represent the time-variation in the detector orientation. Both can be considered as periodic functions with a period of a year, which allows for an additional rewriting in terms of Fourier series instead of Fourier transforms, as will be developed in Sec.~\ref{}.

>>>>

\newcommand{\SM}[1]{{\color{Red} #1}}
\newcommand{\jgb}[1]{{\color{DarkGreen} #1}}

@@

\author{Sylvain Marsat}
\affiliation{Max Planck Institute for Gravitational Physics (Albert Einstein Institute), Am M\"uhlenberg 1, Potsdam-Golm, 14476, Germany}
\author{John G. Baker}
\affiliation{Gravitational Astrophysics Laboratory, NASA Goddard Space Flight Center, 8800 Greenbelt Rd., Greenbelt, MD 20771, USA}
\affiliation{Joint Space-Science Institute, University of Maryland, College Park, MD 20742, USA}

@@

We present a Fourier-domain based approach modulations and delays of gravitational wave signals, a problem which arises in two different contexts. For space-based detectors like LISA, the orbital motion of the detector introduces a time-dependency in the response of the detector, consisting of both a modulation and a varying delay. In the context of signals from precessing spinning binary systems, a useful tool for building models of the waveform consists in representing the signal as a time-dependent rotation of a quasi-non-precessing waveform. In both cases, being able to compute transfer functions for these effects directly in the Fourier domain may enable performance improvements for data analysis applications by using fast frequency-domain waveforms. Our results generalize previous approaches based on the stationary phase approximation, extending them by including delays and higher-order corrections while being applicable to a broader class of signals including full inspiral-merger-ringdown waveforms. In the LISA case, we find that higher-order corrections remain small except at the two extremes of the detector's frequency band, where we show that including them reduces the errors. In the case of precessing binaries, we perform a limited exploration of the merger-ringdown range based on a toy model for the precession. We find that, depending on the post-merger frame velocity, higher-order corrections can become quantitatively large to the point of challenging our perturbative formalism, while affecting mainly the amplitude and giving rise to limited unfaithfulness. We further present an alternative direct convolution approach to accurately represent these post-merger features.

@@

With the recent detections of gravitational-wave signals emitted by the coalescences of binary black holes, announced by the LVC collaboration during the first and second obervation runs of the two LIGO instruments~\cite{LIGO-theevent-2016,LIGO-christmasevent-2016,LIGO-O1BBH-2016}\jgb{[update refs]}, gravitational-wave astronomy has entered its observational era. With the conclusion of LIGO's second observation run, and the recent extension of the ground-based detectors network with Virgo~\cite{Virgo} ( and later also KAGRA~\cite{KAGRA} and LIGO-India~\cite{INDIGO}), many more detections from compact object coalescences are to come, with binary black holes being expected to be the dominant sources~\cite{} \SM{[C]}.

Moreover, the European Space Agency has issued a mission call for its L3 space mission with a tentative launch around 2034, adressing the science theme ``The Gravitational Universe'', possibly opening the way for a space-borne observatory based on the concept of the Laser Interferometer Space Antenna (LISA)~\cite{LISA09,LISA17}. Among other sources of gravitational waves, LISA will be able to detect and characterize comparable-mass binary black hole coalescences to cosmological distances over a wide range of masses, from high-redshift observations of supermassive black hole binaries with $M\sim 10^{7} \Msol$ down to the observation of LIGO-type sources with $M\sim 10^{2} \Msol$~\cite{Sesana16}.

Data analysis for gravitational-wave observations of compact binary coalescences require accurate models (or templates) for the signals, both for ensuring efficient detections of signals that may be buried in instrumental noise, and to extract the physical parameters of the source in a subsequent analysis. The the greater sensitivity of LISA and other future instruments will require further increases in the accuracy of computationally efficient signal templates. This motivates an important and ongoing effort to improve gravitational wave templates. Two important focus areas for template improvements are need to cover all phases of the evolution of the system, including the inspiral, merger and ringdown (IMR), which became possible thanks to the breakthrough of numerical relativity~\cite{Pretorius05,Baker+06,Campanelli+06}, and the inclusion of the effects of the spins of the compact objects, that can lead to a precession of the orbital plane~\cite{ACST94,K95}. Besides accuracy, the computational efficiency of template generation is crucial. Bayesian analysis for parameter estimation of gravitational-wave signals, as was performed for the LIGO detections~\cite{LIGO-theeventPE-2016,LIGO-O1BBH-2016}, may require millions of evaluations of the likelihood function to sample the posterior probability distribution.

The GW community is making progress assembling higher-fidelity tools for these coming challenges.
State-of-the-art IMR templates combine information from the perturbative results of post-Newtonian theory covering the inspiral (see e.g.~\cite{BlanchetLiving}) and from numerical relativity simulations covering the end of the inspiral and the merger-ringdown phase (see e.g.~\cite{Pfeiffer12}). The Effective-One-Body (EOB) approach~\cite{BD99} does so by incorporating a resummation of post-Newtonian results and a calibration to numerical relativity, and has been extended to include both spins of the two bodies~\cite{Taracchini+13, Pan+13, Bohe+16}. Phenomenological templates extend post-Newtonian models by postulating an analytic ansatz for the Fourier-domain amplitude and phase, fitting coefficients against numerical relativity waveforms~\cite{Husa+15, Khan+15}, and can include a simplified model for precession~\cite{Hannam+13}. The second type of models produce directly the amplitude and the phase of the Fourier-domain waveform components in the form of a piecewise analytic function. EOB waveforms, by contrast, require integration of ordinary differential equations in the time-domain, which can be time-consuming. For the non-precessing case, Reduced Order Models (ROM), also called surrogate models, have been developed to speed up considerably the waveform generation, without losing accuracy~\cite{Field+13, Puerrer14, Bohe+16}. Both approaches are therefore able to efficiently produce IMR Fourier-domain waveforms. An important aspect is that the resulting waveforms can be represented by an amplitude and phase for each mode, with only a few hundred samples~\cite{Puerrer14}.

A complete representation of spin effects in fast IMR templates still remains a frontier of gravitational wave signal modelling. In presence of misaligned spins, the system will endure precession of the orbital plane as it evolves, leading to modulations of the signal as seen by the observer~\cite{ACST94}. Precessing EOB templates incorporate all six parameters of the two spins~\cite{Pan+13, BTB16} but still require a costly time-domain integration; while precessing phenomenological templates~\cite{Hannam+13} decrease the dimensionnality by using effective spin parameters representing only the dominant effects of the spins. To model the effect of precession on the emitted waveform, it has been proposed~\cite{BCV03b, BCPTV05, Schmidt+10, OShaughnessy+11, Boyle+11} to decompose precessing waveforms by performing a time-dependent rotation, following the precession of the orbital plane. The resulting waveform in the rotated frame can then be modelled by a non-precessing waveform, an approximation which is used both in the construction of the inspiral part of precessing EOB waveforms~\cite{Pan+13} and in the construction of precessing phenomenological waveforms~\cite{Hannam+13}. To follow this modelling approach and efficiently create Fourier-domain waveforms, one needs to understand how to translate the time-domain modulations created by the frame rotation into a Fourier-domain transfer function, which is our first motivation.

Prospective parameter recovery studies for LISA observations of comparable-mass binary black hole coalescences aim at characterizing the scientific performance of the future experiment. Although LISA\ref{LISAMissionWP} is still at least a decade from launch, the ESA L3 mission is rapidly progressing into its early stages where mission design choice must be informed by science impact analysis. However, the recent progresses in waveform modelling, most notably the availability of cheap IMR waveforms, have not yet made their way to the field of LISA data analysis studies. In the past, such studies used mainly post-Newtonian, inspiral-only models, and were limited to the Fisher-matrix formalism instead of full Bayesian analysis (among numerous works, see e.g. \cite{Cutler97,LISAPE09}). Refs.~\cite{McWilliams+09,McWilliams+10,McWilliams+11} explored the effects of including the merger and ringdown, keeping the mass parameters fixed. However, the Fisher matrix analysis, although considerably cheaper to perform, is valid only in the limit of a high signal to noise ratio (SNR) and has known shortcomings~\cite{Vallisneri08}. Several implementations of Bayesian analyses were developed~\cite{CP06,Wickham+06,Babak+08,GP09,Feroz+09,Petiteau+10,PC13} (see also~\cite{MLDC4} and references therein), with Ref.~\cite{PC15} proposing a direct comparison of Fisher and Bayes analyses, but they were limited to inspiral-only waveforms. One exception was Ref.~\cite{Babak+08} where IMR signals were used in a Bayesian study, but there the masses of the black holes were kept fixed. In the recent Ref.~\cite{Klein+15}, the limitation to inspiral-only precessing post-Newtonian waveforms was compensated by an SNR reweighting procedure. Thus, due mainly to their high computational cost, full-fledged Bayesian parameter estimation simulations for IMR signals are still missing. \jgb{[Maybe there is too much detail in this paragraph?]}

In contrast with current ground-based detectors, which are typically sensitive to chirping binaries for less than a second, LISA signals may accumulating over months or years.
The response of a LISA-type instrument is thus time-dependent~\cite{Cutler97}.
The motion and change of orientation of the detector constellation along its orbit lead to significant time variability in the form of a modulation and a varying delay. These effeccts then convey information about the localization of the gravitational-wave source in the sky. The complicated detector response increase the computational cost of parameter-estimation analyses. To leverage the performance of state-of-the-art Fourier-domain IMR templates~\cite{BTB16,Khan+15}, one needs to be able to process the signals through the time-dependent response of the dectector while staying in the Fourier domain.
%, which is our second motivation.

The purpose of this paper is to introduce a formalism for efficiently processing signals through a time-domain modulation and delay within the Fourier domain, while retaining the compactness of a Fourier-domain amplitude and phase representation of the signals. This will allow us to address both the issue of the Fourier-domain response of the LISA instrument, as well as the issue of Fourier-domain precession modulation for IMR signals from precessing binaries.

@@

The plan of the paper is as follows. In Sec.~\ref{sec:motivation}, \sout{we will introduce our notations and conventions for Fourier-domain waveforms}, motivate the problem of Fourier-domain modulation and introduce the relevant timescales for both the response of LISA-type detectors and the modulation of precessing signals. In Sec.~\ref{sec:formalism}, we present our general formalism, give its leading order approximation as well as higher-order corrections, introduce new timescales based on the Fourier-domain signal, refine the previous results for both the quadratic-in-phase corrections and the treatment of the delays, and make the connection to previous approaches of the problem. We then apply our formalism to the response of the LISA detector in Sec.~\ref{sec:LISA}, and to the case of signals from precessing binaries in Sec~\ref{sec:precession}. We discuss and summarize our results in Sec.~\ref{sec:discussion}.

@@

\subsection{Basic considerations}
\label{subsec:basic}

We can jointly consider challenges in the contexts of both the LISA response and GW precession applications in terms of a more general problem which covers both domains. In each case must determine the Fourier transform of a signal $h(t)$ to which we apply a varying delay $d(t)$ before multiplication by a modulation function $F(t)$,
\be
\label{eq:delay-mod-defs}
	h_{d}(t) = h(t+d(t)) \,, \quad s(t) = F(t)h_{d}(t) \,,
\ee
and seek an efficient way to commpute the Fourier transform $\tilde{s}(f)$, expressed by means of a Fourier-domain transfer function $\calT$ such that
\be\label{eq:deftransfer}
	\tilde{s}(f) \equiv \calT(f) \tilde{h}(f) \,. 
\ee

In the case of precessing binaries, the delays will be absent and the modulation functions will be the time-dependent Wigner coefficients applied to rotate the waveform from a precessing frame, where it verifies a restored mode hierarchy, to an inertial frame where the observations take place. In the case of a LISA-type detector, the signal $h(t)$ will simply be the waveform in a fixed heliocentric frame, and the delays will come from the motion of each detector in the wave front, while the modulation will represent the time-variation in the detector orientation. \jgb{MOVE:Both can be considered as periodic functions with a period of a year, which allows for an additional rewriting in terms of Fourier series instead of Fourier transforms, as will be developed in Sec.~\ref{}.NOTE: Probably move other technical partso f the par. as well.}

\jgb{In full generality computing the transfer function $\calT(f)$ will require a convolution, a costly integral (numerically a sum) over the full frequency domain for each value of $f$. For our context though, and perhaps for other applications, there are some properties of $h(t)$, $d(t)$ and $F(t)$ which we can exploit for a more efficient computation.}

\jgb{First consider the waveforms. There is a separation between the timescales of orbital motion and radiation-reaction. This leads to a general feature of many GW signals, including binary inspirals and black hole mergers, that the signal is relatively localized in time-frequency.  As far as $h(f)$ is concerned then, for given $f$, only a limited range of times should be relevant in $d(t)$ and $F(t)$ so that we may expect to find a more efficient treatment for the transfer function.}

\jgb{As is well known (refs) this is especially true for non-precessing systems when the signal is decomposed in spin-weighted spherical harmonic components with respect to the source orientation angles.  In amplitude-phase form the individual spherical harmonic components of even long-lasting signals may be treated as smoothly varying functions of frequency.  An important consequence is that these signal components can be accurately represented in terms of a relatively coarsely sampled frequency grid, but we will also use this property in transfer function treatment.  (See Appendix~\ref{sec:notation} for details of our notation and conventions for Fourier transforms and spherical harmonics.)}

\jgb{Next consider the delay and modulation functions.  Again we can identify a separation of time scales between the modulation or delay and the gravitational-wave period. This is either the precession timescale or for the LISA response the annual-orbital motion timescale for the constellation.}

\jgb{Our objective is to find an efficient approximate treatment for $\calT(f)$ which allows us to exploit these properties in a way which is extensible to the high-accuracies which will be required by LISA and other future gravitational-wave instruments, yet which widely applicable to GW analysis, without unnecessary limiting assumptions. Before introducing developing our formalism, though, we review the salient features of our application problems.}

@@

The response of a detector of the LISA type to an incident gravitational wave can be written in two different, equivalent forms, in terms of phase or frequency measurements. Here we will work with the second representation, which will prove more convenient for our purposes. Moreover, various notation and conventions have been used in the literature to label the spacecrafts and describe their orbits. We refer the reader to~\cite{Vallisneri04} for a comparative account on these various conventions. In this work, we will use the conventions of~\cite{Vallisneri04}, which were also used in the Mock LISA Data Challenges (MLDC)~\cite{} \SM{[C]}.

@@

As a preliminary step, we recall here an approximation widely used for similar purposes in treating gravitational waves signals emitted by inspiralling binaries, the Stationary Phase Approximation (thereafter SPA). It applies in general to chirping signals, and we refer the reader to~\ref{} for details. One first writes the time-domain signal in amplitude and phase form as $h(t) = a(t) e^{-2i\varphi(t)}$, where for gravitational wave signals $\varphi$ will correspond to the orbital phase, with the orbital frequency being $\omega = \dot{\varphi}$. In order to keep close to the notation used in the gravitational wave literature, we introduced a factor of 2 in the phase of the wave, which is appropriate for the dominant 22 harmonic of the signal. The approximation then applies to signals verifying the conditions \SM{[say that the third is sometimes missing ?]}
\be\label{eq:conditionSPA}
	\left| \frac{\dot{a}/a}{\omega} \right| \ll 1\,, \quad \left|\frac{\dot{\omega}}{\omega^{2}} \right| \ll 1\,, \quad \left| \frac{(\dot{a}/a)^{2}}{\dot{\omega}} \right| \ll 1 \,.
\ee
Since the integral~\eqref{eq:defFT} defining the Fourier transform is rapidly oscillatory unless the term $2\pi f t$ cancels the evolution of $-2\varphi(t)$, its support is well centered around the point of stationary phase. This determines the time-to-frequency correspondence in the stationary phase approximation, and leads to the definition of this time as an implicit function of frequency by the relation
\be\label{eq:deftfSPA}
	\omega(t_{f}^{\rm SPA}) = \pi  f \,.
\ee
For a chirping signal of increasing phase, $\omega>0$ and $\dot{\omega}>0$, there is a unique point of stationary phase, located in the positive frequency range $f>0$. Using the conditions above, one can formally expand the signal around $t_{f}^{\rm SPA}$ to quadratic order in time, according to
\begin{align}
	\tilde{h}_{\rm SPA} (f) &\simeq a(t_{f}^{\rm SPA}) \exp\left[2i\pi f t_{f}^{\rm SPA}-2i\varphi(t_{f}^{\rm SPA}) \right] \nn\\
	& \qquad \cdot \int \ud t \, e^{-i \dot{\omega} (t_{f}^{\rm SPA}) (t-t_{f}^{\rm SPA})^{2}} \,.
\end{align}
Note that to be able to treat the amplitude as a constant in the integral above, we used the third condition in~\eqref{eq:conditionSPA}. The resulting complex Gaussian integral yields\footnote{Note that the expressions given here are valid for the $22$ mode of the waveform.}
\begin{subequations}
\begin{align}
	\tilde{h}_{\rm SPA}(f) &= A_{\rm SPA}(f) e^{-i\Psi_{\rm SPA}(f)} \,, \\
	A_{\rm SPA}(f) &= a(t_{f}^{\rm SPA}) \sqrt{\frac{\pi}{\dot{\omega}(t_{f}^{\rm SPA})}} \,, \\
	\Psi_{\rm SPA}(f) &= 2\varphi(t_{f}^{\rm SPA}) - 2\pi f t_{f}^{\rm SPA} + \frac{\pi}{4} \,.
\end{align}
\end{subequations}
Ref.~\cite{Droz+99} evaluated the first correction to this approximation, within the context of post-Newtonian signals, and found that it can be considered as a term of the fifth post-Newtonian order, beyond the accuracy level of our current best models~\cite{BlanchetLiving}.

\jgb{In the Newtonian, low-frequency limit each of the combinations \eqref{eq:conditionsSPA} assumed to be small in the SPA limit indeed vanishes at $\calO{((v/c)^5)}$, but as the system approaches merger, this condition is no-longer satisfied.  Near-merger however the treatment is not applicable, so that this is not a viable approach for binary systems beyond insprial. Nonetheless the SPA, has been a useful workhorse in many gravitational analysis involving frequency domain transfer functions of the form \eqref{eq:deftransfer}.  The usual approach is simply to replace any time dependencies appearing the transfer function using \eqref{eq:deftfSPA}.  We will later return to further consider the SPA treatment in the context of a more general formalism.}

@@

\jgb{In this section we develop a Fourier-domain formalism for treating time-delays and temporally multiplicative signal transformations, where the period of primary signal oscilations is short compared to the timescale on which the modulating functions vary.}

@@

\jgb{We begin by expression our delay and modulation function definitions in \eqref{eq:delay-mod-defs} using the Fourier transform (note conventions in \eqref{eq:defFT})},
\begin{align}
h_{d}(t) & =  h(t+d(t)) \nonumber\\
&=\int \ud f \, e^{-2i\pi f (t+d(t))}\tilde{h}(f) \,,
\end{align}
and
\begin{align}
  \tilde{s}(f) &= F(t)h_{d}(t) \nn \\
  &= \int \ud t \, e^{2i\pi f t} F(t)  \int \ud f' \, e^{-2i\pi f' (t+d(t))}\tilde{h}(f') \nn\\
	&= \int \ud f' \, \tilde{h}(f-f') \int \ud t \, e^{2i\pi f' t} e^{-2i\pi (f-f') d(t)} F(t) \,.
\end{align}
The last equation can then be rewritten as a kind of convolution with a frequency-dependent Kernel, according to
\be\label{eq:FDkernel}
	\tilde{s}(f) = \int \ud f' \, \tilde{h}(f-f') \tilde{G}(f-f',f') \,,
\ee
where we introduced the frequency-dependent function of time $G(f,t)$, and its Fourier transform in the auxiliary frequency $f'$, denoted by $\tilde{G}(f,f')$, as
\begin{subequations}\label{eq:defg}
\begin{align}
	G(f,t) &= e^{-2i\pi f d(t)} F(t) \,, \\
	\tilde{G}(f,f') &= \int \ud t \, e^{2i\pi f' t} e^{-2i\pi f d(t)} F(t) \,.
\label{eq:defGtilde}
\end{align}
\end{subequations}
In the absence of delays $d(t)$, as in the case of precessing binaries, the function $G$ loses its frequency-dependence and becomes a modulation in the form of a function of time $F(t)$, and the result~\eqref{eq:FDkernel} above reduces to the usual convolution theorem for the Fourier transform.

\jgb{A direct computation of \eqref{eq:FDkernel} using \eqref{eq:defGtilde} will generally be computationally demanding, but when $\tilde{G}(f,f')$ has only limited support over $f'$, we can seek an accurate but more efficient approximation.} In our approach, two assumptions will be crucial for our approximations to be valid:
\begin{itemize}
	\item The time scales entering the signal $h(t)$ are well separated from the ones entering the modulation and the delay, so that $\tilde{G}(f-f',f')$ has support limited to $|f'| < f'_{\rm max}$, with $f'_{\rm max}$ small compared the characteristic scale for the variation of the Fourier-domain signal;
	\item The original signal has a smooth amplitude and phase in the Fourier-domain, slowly varying over the scale $f'_{\rm max}$, so that a local treatment in $f'$ of the convolution in~\eqref{eq:FDkernel} will be possible.
\end{itemize}
Although well verified for the case of LISA-type detectors, the first assumption can be a limitation of the approximation in the case of preccesing binaries, and we will carefully explore its validity in Sec.~\ref{}.
